Towards a framework for reasoning about stateful compiler IRs using coinductive reasoning.

### 2024-06-25 paper.tex

Now, formal verification has been done before. Most notable is the
CompCert project: a fully verified end-to-end C compiler~\cite{leroyCompCertFormallyVerified}. Yet, the vast majority of
code today is still being compiled with non-verified compilers. We ask,
thus, how do we formally verify not just \emph{a} compiler, but, how do we
scale formal verification to \emph{mainstream} compilers.

@@

We need the full power of a theorem prover to be able to capture the
full extensible nature of MLIR. Still, to recover the ease-of-use that
SMT-solver-based tools provide, we've also been investing in building
solid automation for bitvector proofs in collaboration with the Lean
community.

In the next two years, I'd like to look into a feature of MLIR called
\emph{Graph Regions}. Specific dialects may use such regions to allow a
variable to be used \emph{before} it is defined --- hence, turning the
def-use chain into a general graph. Generally, the semantics of such
dialects are statefull, and cycles are broken by looking at the value of
a variable at a previous state. For example, graph regions are used by
dialects that model hardware,
where variables model wires and cycles represent physical cycles in the
circuit~\cite{eldridgeMLIRHardwareCompiler}.

@@

Finally, we return to the main objective: formalizing compiler IRs.

\begin{quote}
Apr '25 - June '25
\end{quote}

Therefore, I'd like to use my library for coinductive types to define a
coinductive monad which supports computable infinite behaviours, drawing
inspiration from the interaction tree library for coq.
\cite{xiaInteractionTreesRepresenting2020} This addresses a
weakness in LeanMLIR while simultaneously giving a nice case study for
the usability of the coinduction library.

\begin{quote}
June '25 - Sep '25
\end{quote}

Here, \texttt{y} is obtained as the bitwise and of the argument
\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
sense, but this style of definition is common in hardware
representations, where this is interpreted as: ``the value of \texttt{y}
in the \emph{current} clock cycle is computed by taking the bitwise and
of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''.~\footnote{
  This example is simplified, in reality you'd have to specify a
  register instruction to control when a value is stored across clock
  cycles. See https://circt.llvm.org/ for more details}
To formally model this, we could denote variables as infinite streams of bitvectors,
which gives us the value at every possible future clock cycle. Another
example of a dialect with graph regions is the FSM, or Finite State
Machine, dialect.

% \subsection{Stretch Goal: Iris}\label{stretch-goal-iris}

% \begin{quote}
% {[}!TODO{]} Talk about how denotational semantics are cool, and how the
% aforementioned improvements will let us define dialects with
% side-effects, but reasoning about them will still be a bit painful. We'd
% really like to reason with program logics instead, in particular, iris:
% \cite{jungIrisMonoidsInvariants2015}\cite{jungHigherorderGhostState2016}
% \end{quote}

>>>>

Now, formally verified compilation has been done before.
Most notable is the CompCert project: a fully verified end-to-end C compiler~\cite{leroyCompCertFormallyVerified}.
Yet, the vast majority of
code today is still being compiled with non-verified compilers. We ask,
thus, how do we formally verify not just \emph{a} compiler, but, how do we
scale formal verification to \emph{mainstream} compilers.

@@

We need the full power of a theorem prover to be able to capture the
extensible and high-level nature of MLIR dialects.
Still, to recover the ease-of-use that SMT-solver-based tools provide, 
we've also been investing in building solid automation for bitvector proofs in collaboration with the Lean community.

In the next two years, I'd like to look into a feature of MLIR called
\emph{Graph Regions}. Specific dialects may use such regions to allow a
variable to be used \emph{before} it is defined --- hence, turning the
def-use chain into a general graph. Generally, the semantics of such
dialects are stateful, and cycles are broken by looking at the value of
a variable at a previous state.
For example, graph regions are used by
dialects that model hardware,
where variables model wires and cycles represent physical cycles in the
circuit~\cite{eldridgeMLIRHardwareCompiler}.

@@

% \subsection{Stretch Goal: Iris}\label{stretch-goal-iris}
% 
% \begin{quote}
% {[}!TODO{]} Talk about how denotational semantics are cool, and how the
% aforementioned improvements will let us define dialects with
% side-effects, but reasoning about them will still be a bit painful. We'd
% really like to reason with program logics instead, in particular, iris:
% \cite{jungIrisMonoidsInvariants2015}\cite{jungHigherorderGhostState2016}
% \end{quote}
